---
title: "Multi-contrast enrichment analysis of DUX data"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
theme: cosmo
---

# Introduction

The goal here is to perform enrichment analysis with the human and mouse DUX data.

```{r,libs}
library("mitch")
library("DESeq2")
library("edgeR")

```


# Prepare Jones data

1. MCM: strain: ACTA1-MCM/+; treatment: no tamoxifen (CONTROL)

2. FLExD: strain: FLExDUX4/+treatment: no tamoxifen (DO NOT USE)

3. dTGM: strain: FLExDUX4/+;MCM/+ treatment: one 5 mg/kg IP injection of tamoxifen (CASE - MILD)

4. dTGS: FLExDUX4/+;MCM/+ treatment: two 10 mg/kg IP injections of tamoxifen on consecutive days (CASE - SEVERE)

```{r,jones}

# mouse GSE122562
#URL="https://ftp.ncbi.nlm.nih.gov/geo/series/GSE122nnn/GSE122562/suppl/GSE122562_gene_counts$
j <- read.table("GSE122562_gene_counts_table.tsv.gz",header=TRUE,row.names=1)
j[,1]<-NULL
ss <- as.data.frame(colnames(j))
ss$mcm <- grepl("MCM",ss[,1])*1
ss$flexd <- grepl("FLExD",ss[,1])*1
ss$dtgm <- grepl("dTGM",ss[,1])*1
ss$dtgs <- grepl("dTGS",ss[,1])*1
rownames(ss) <- ss[,1]
ss[,1]<-NULL
# here set the contrast
sss <- subset(ss,mcm==1|dtgm==1)
design <- model.matrix(~sss$dtgm)
rownames(design) <- rownames(sss)
jj <- j[,which(colnames(j) %in% rownames(design) )]
jj <- jj[which(rowMeans(jj)>10),]
z <- DGEList(counts=jj)
z <- calcNormFactors(z)
z <- estimateDisp(z, design,robust=TRUE,prior.df=1)
fit <- glmFit(z, design)
lrt<-glmLRT(fit)
dge<-as.data.frame(topTags(lrt,n=Inf))
dge$dispersion<-lrt$dispersion
dge<-merge(dge,lrt$fitted.values,by='row.names')
rownames(dge)=dge$Row.names
dge$Row.names=NULL
dge<-dge[order(dge$PValue),]
head(dge,10)
jm <- dge

sss <- subset(ss,mcm==1|dtgs==1)
design <- model.matrix(~sss$dtgs)
rownames(design) <- rownames(sss)
jj <- j[,which(colnames(j) %in% rownames(design) )]
jj <- jj[which(rowMeans(jj)>10),]
z <- DGEList(counts=jj)
z <- calcNormFactors(z)
z <- estimateDisp(z, design,robust=TRUE,prior.df=1)
fit <- glmFit(z, design)
lrt<-glmLRT(fit)
dge<-as.data.frame(topTags(lrt,n=Inf))
dge$dispersion<-lrt$dispersion
dge<-merge(dge,lrt$fitted.values,by='row.names')
rownames(dge)=dge$Row.names
dge$Row.names=NULL
dge<-dge[order(dge$PValue),]
head(dge,10)
js <- dge

```

# Read in Watt data

```{r,read}

# human 
h <- read.table("dux2_edgeR_edger.tsv",header=TRUE,row.names=1)
rownames(h) <- sapply(strsplit(rownames(h),"_"),"[[",1)

# mouse simple
# we're not using the factorial
ms <- read.table("EVvsDUX_wDOX_paired.tsv",header=TRUE,row.names=1)
rownames(ms) <- sapply(strsplit(rownames(ms),"_"),"[[",1)

# ortholog mapping
orth <- read.table("mouse2human.txt.sort")

# gene name mapping human
gth <- orth[,c(1,3)]

# gene name mapping human
gtm <- orth[,c(2,3)]

```

# Get gene sets

```{r,geneset}

#download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", destfile="ReactomePathways.gmt.zip")
#unzip("ReactomePathways.gmt.zip")
genesets <- gmt_import("ReactomePathways.gmt")

```

# Prepare data for running mitch

```{r,mitch1}

hh <- mitch_import(x=h,DEtype="edger",geneTable=gth)

mm <- list("ms"=ms,"jm"=jm,"js"=js)
mmm <- mitch_import(x=mm,DEtype="edger",geneTable=gtm)


mg <- merge(hh,mmm,by=0)
rownames(mg) <- mg[,1]
mg[,1]<-NULL
colnames(mg) <- c("human","mouse","jmod","jsev")
dim(mg)
head(mg)

res1 <- mitch_calc(x=mg,genesets=genesets,priority="effect")
head(res1$enrichment_result,20)
unlink("mitch_plots_simple_eff.html")
capture.output(
    mitch_report(res1,"mitch_plots_simple_eff.html")
  ,file = "/dev/null", append = FALSE,
  type = c("output", "message"), split = FALSE)

#res2 <- mitch_calc(x=mg,genesets=genesets,priority="significance")
#head(res2$enrichment_result,20)
#unlink("mitch_plots_simple_sig.html")
#mitch_report(res2,"mitch_plots_simple_sig.html")

```

# Session information

```{r,session}

sessionInfo()

```
