---
title: "Multi-contrast enrichment analysis of DUX data"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
theme: cosmo
---

Source: https://github.com/markziemann/dux

# Introduction

The goal here is to perform enrichment analysis with the human and mouse DUX data.

```{r,libs}

suppressPackageStartupMessages({
  library("mitch")
  library("DESeq2")
  library("edgeR")
  library("eulerr")
  library("UpSetR")
  library("gplots")
  library("getDEE2")
})

```

# Prepare human FSHD data

```{r,fshd}

ss <- read.table("fshd.txt",header=TRUE)
md <- getDEE2Metadata("hsapiens")
mymd <- md[which(md$Sample_name %in% ss$Accession),]
ss <- merge(mymd,ss,by.x="Sample_name",by.y="Accession")
SRRvec <- ss$SRR_accession
x <- getDEE2(species="hsapiens",SRRvec=SRRvec,legacy=TRUE,metadata=md)
xx <- x$GeneCounts

design <- model.matrix(~ss$Group)
rownames(design) <- ss$SRR_accession

xx <- xx[,which(colnames(xx) %in% rownames(design) )]
xx <- xx[which(rowMeans(xx)>10),]
z <- DGEList(counts=xx)
z <- calcNormFactors(z)
z <- estimateDisp(z, design,robust=TRUE,prior.df=1)
fit <- glmFit(z, design)
lrt <- glmLRT(fit)
fshd <- as.data.frame(topTags(lrt,n=Inf))
write.table(fshd,file="fshd.tsv",sep='\t',quote=FALSE)

fshd_up <- rownames(subset(fshd,FDR<0.05 & logFC>0))
fshd_dn <- rownames(subset(fshd,FDR<0.05 & logFC<0))
fshd_subset <- rownames(subset(fshd,FDR<0.05 & logFC>2))

# Read in Watt data
# human 
h <- read.table("dux2_edgeR_edger.tsv",header=TRUE,row.names=1)
rownames(h) <- sapply(strsplit(rownames(h),"_"),"[[",1)

h_up <- rownames(subset(h,FDR<0.05 & logFC>0))
h_dn <- rownames(subset(h,FDR<0.05 & logFC<0))
h_subset <- rownames(subset(h,FDR<0.05 & logFC>2))

v1 <- list("fshd up"=fshd_up, "fshd dn"=fshd_dn,
  "h up"=h_up,"h dn"=h_dn)

plot(euler(v1),quantities = TRUE)

```

# Prepare Jones data

1. MCM: strain: ACTA1-MCM/+; treatment: no tamoxifen (CONTROL)

2. FLExD: strain: FLExDUX4/+treatment: no tamoxifen (DO NOT USE)

3. dTGM: strain: FLExDUX4/+;MCM/+ treatment: one 5 mg/kg IP injection of tamoxifen (CASE - MILD)

4. dTGS: FLExDUX4/+;MCM/+ treatment: two 10 mg/kg IP injections of tamoxifen on consecutive days (CASE - SEVERE)

```{r,jones}

# mouse GSE122562
#URL="https://ftp.ncbi.nlm.nih.gov/geo/series/GSE122nnn/GSE122562/suppl/GSE122562_gene_counts$
j <- read.table("GSE122562_gene_counts_table.tsv.gz",header=TRUE,row.names=1)
j[,1]<-NULL
ss <- as.data.frame(colnames(j))
ss$mcm <- grepl("MCM",ss[,1])*1
ss$flexd <- grepl("FLExD",ss[,1])*1
ss$dtgm <- grepl("dTGM",ss[,1])*1
ss$dtgs <- grepl("dTGS",ss[,1])*1
rownames(ss) <- ss[,1]
ss[,1]<-NULL
# here set the contrast
sss <- subset(ss,mcm==1|dtgm==1)
design <- model.matrix(~sss$dtgm)
rownames(design) <- rownames(sss)
jj <- j[,which(colnames(j) %in% rownames(design) )]
jj <- jj[which(rowMeans(jj)>10),]
z <- DGEList(counts=jj)
z <- calcNormFactors(z)
z <- estimateDisp(z, design,robust=TRUE,prior.df=1)
fit <- glmFit(z, design)
lrt<-glmLRT(fit)
dge<-as.data.frame(topTags(lrt,n=Inf))
dge$dispersion<-lrt$dispersion
dge<-merge(dge,lrt$fitted.values,by='row.names')
rownames(dge)=dge$Row.names
dge$Row.names=NULL
dge<-dge[order(dge$PValue),]
head(dge,10)
jm <- dge
write.table(jm,file="jm.tsv",sep='\t',quote=FALSE)

sss <- subset(ss,mcm==1|dtgs==1)
design <- model.matrix(~sss$dtgs)
rownames(design) <- rownames(sss)
jj <- j[,which(colnames(j) %in% rownames(design) )]
jj <- jj[which(rowMeans(jj)>10),]
z <- DGEList(counts=jj)
z <- calcNormFactors(z)
z <- estimateDisp(z, design,robust=TRUE,prior.df=1)
fit <- glmFit(z, design)
lrt<-glmLRT(fit)
dge<-as.data.frame(topTags(lrt,n=Inf))
dge$dispersion<-lrt$dispersion
dge<-merge(dge,lrt$fitted.values,by='row.names')
rownames(dge)=dge$Row.names
dge$Row.names=NULL
dge<-dge[order(dge$PValue),]
head(dge,10)
js <- dge

```

Now to see how similar these contrasts are.
The overlap is really high, therefore I suggest only using the moderate contrast in downstream mitch analysis.

```{r,jvenn}

jm_up <- rownames(subset(jm,FDR<0.05 & logFC>0))
jm_dn <- rownames(subset(jm,FDR<0.05 & logFC<0))

js_up <- rownames(subset(js,FDR<0.05 & logFC>0))
js_dn <- rownames(subset(js,FDR<0.05 & logFC<0))

v1 <- list("mod up"=jm_up, "mod dn"=jm_dn,
  "sev up"=js_up,"sev dn"=js_dn)

plot(euler(v1),quantities = TRUE)

```

# Read in Watt data

```{r,read_m}

# mouse simple
# we're not using the factorial
ms <- read.table("EVvsDUX_wDOX_paired.tsv",header=TRUE,row.names=1)
rownames(ms) <- sapply(strsplit(rownames(ms),"_"),"[[",1)

# ortholog mapping
orth <- read.table("mouse2human.txt.sort")

# gene name mapping human
gth <- orth[,c(1,3)]

# gene name mapping human
gtm <- orth[,c(2,3)]

```

Calculate overlap

```{r,ol}

jm_up <- rownames(subset(jm,FDR<0.05 & logFC>0))
jm_dn <- rownames(subset(jm,FDR<0.05 & logFC<0))
jm_subset <- rownames(subset(jm,FDR<0.05 & logFC>2))

h_up <- rownames(subset(h,FDR<0.05 & logFC>0))
h_dn <- rownames(subset(h,FDR<0.05 & logFC<0))
h_subset <- rownames(subset(h,FDR<0.05 & logFC>2))

ms_up <- rownames(subset(ms,FDR<0.05 & logFC>0))
ms_dn <- rownames(subset(ms,FDR<0.05 & logFC<0))
ms_subset <- rownames(subset(ms,FDR<0.05 & logFC>2))

l <- list("jm_up"=jm_up,"jm_dn"=jm_dn,
  "h_up"=h_up,"h_dn"=h_dn,
  "fshd up"=fshd_up,"fshd dn"=fshd_dn,
  "ms_up"=ms_up,"ms_dn"=ms_dn)

barplot(unlist(lapply(l,length)),main='no. DEGs')

# here are the values
unlist(lapply(l,length))

# convert lists to human gene names
jm_up <- unique(gtm[which(gtm$V2 %in% jm_up ),2])
jm_dn <- unique(gtm[which(gtm$V2 %in% jm_dn ),2])
jm_subset <- unique(gtm[which(gtm$V2 %in% jm_subset ),2])
writeLines(jm_subset,con="jonesmouse_subset.txt")

h_up <- unique(gth[which(gth$V1 %in% h_up ),2])
h_dn <- unique(gth[which(gth$V1 %in% h_dn ),2])
h_subset <- unique(gth[which(gth$V1 %in% h_subset ),2])
writeLines(h_subset,con="humanmodel_subset.txt")

ms_up <- unique(gtm[which(gtm$V2 %in% ms_up ),2])
ms_dn <- unique(gtm[which(gtm$V2 %in% ms_dn ),2])
ms_subset <- unique(gtm[which(gtm$V2 %in% ms_subset ),2])
writeLines(ms_subset,con="wattmouse_subset.txt")

fshd_up <- unique(gth[which(gth$V1 %in% fshd_up ),2])
fshd_dn <- unique(gth[which(gth$V1 %in% fshd_dn ),2])
fshd_subset <- unique(gth[which(gth$V1 %in% fshd_subset ),2])
writeLines(fshd_subset,con="fshd_subset.txt")

# venn of subsets
v1 <- list("Jones mouse"=jm_subset,
  "human model"=h_subset,
  "fshd" = fshd_subset,
  "Watt mouse"=ms_subset )

plot(euler(v1),quantities = TRUE,main="subset genes: FDR<0.05 & LogFC>2")

upset(fromList(v1), order.by = "freq", nsets=6)

v1 <- list("jm up"=jm_up, 
  "jm dn"=jm_dn,
  "h up"=h_up,
  "h dn"=h_dn,
  "fshd up" = fshd_up,
  "fshd dn" = fshd_dn,
  "ms up"=ms_up,
  "ms dn"=ms_dn )

plot(euler(v1),quantities = TRUE,main="FDR genes")

upset(fromList(v1), order.by = "freq", nsets=6)

```

This is a bit confusing, so we will do separate Venns, comparing each model data with the human fshd data.
We can see that the concordance with human is higher with the Watt data as compared to the Jones data.
Here, I define concordance ratio as the number of FDR genes that are commonly dysregulated divided by the number of genes
with discordant expression.


```{r,venn2}

# human model

v1 <- list("fshd up"=fshd_up,
  "fshd dn"=fshd_dn,
  "h up"=h_up,
  "h dn"=h_dn)

plot(euler(v1),quantities = TRUE)

h_concordance <- (length(intersect(fshd_up,h_up))+length(intersect(fshd_dn,h_dn)))  /
  (length(intersect(fshd_up,h_dn))+length(intersect(fshd_dn,h_up)))

h_concordance

# jones mouse model

v1 <- list("fshd up"=fshd_up,
  "fshd dn"=fshd_dn,
  "jm up"=jm_up,
  "jm dn"=jm_dn)

plot(euler(v1),quantities = TRUE)

j_concordance <- (length(intersect(fshd_up,jm_up))+length(intersect(fshd_dn,jm_dn)))  / 
  (length(intersect(fshd_up,jm_dn))+length(intersect(fshd_dn,jm_up)))

j_concordance

# Watt mouse model
v1 <- list( "fshd up"=fshd_up,
  "fshd dn"=fshd_dn,
  "ms up"=ms_up,
  "ms dn"=ms_dn )

plot(euler(v1),quantities = TRUE)

ms_concordance <- (length(intersect(fshd_up,ms_up))+length(intersect(fshd_dn,ms_dn)))  /  
  (length(intersect(fshd_up,ms_dn))+length(intersect(fshd_dn,ms_up)))

ms_concordance

barplot(c(h_concordance,j_concordance,ms_concordance),
  names.arg=c("human model","Jones mouse","Watt mouse"),
  main="concordance ratio")

```

# Get gene sets

```{r,geneset}

#download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", destfile="ReactomePathways.gmt.zip")
#unzip("ReactomePathways.gmt.zip")
genesets <- gmt_import("ReactomePathways.gmt")

```

# Prepare data for running mitch

The mitch_import function creates

```{r,mitch1,fig.width=8,fig.height=8}


hh <- list("fshd"=fshd,"human model"=h)
hhh <- mitch_import(x=hh,DEtype="edger",geneTable=gth)

mm <- list("ms"=ms,"jm"=jm)
mmm <- mitch_import(x=mm,DEtype="edger",geneTable=gtm)

mg <- merge(hhh,mmm,by=0)
rownames(mg) <- mg[,1]
mg[,1]<-NULL
colnames(mg) <- c("FSHD","hu_model","Watt_m","Jones_m")
dim(mg)
head(mg)

mg <- apply(mg,2, function(x) {
  x[which(x==Inf)] <- 300 ;  x[which(x==-Inf)] <- -300   ; return(x) 
} )

# pearson
round(cor(mg,method='p'),3)

# spearman
round(cor(mg,method='s'),3)

colfunc <- colorRampPalette(c("blue","white", "red"))
heatmap.2(cor(mg,method='s'),
  cellnote=round(cor(mg,method='s'),3),
  trace="none",scale="none",margins=c(15,15),
  col=colfunc(25), notecex=2 , main="Spearman")

```

Here we run mitch unidimensional and then create Venn diagrams of the FDR gene sets.

```{r,mitch_unidimensional,fig.height=8,fig.width=8}

fshdm <- mitch_import( fshd,DEtype="edger",geneTable=gth )
fshdmres <- mitch_calc(x=fshdm,genesets=genesets,priority="effect",resrows=100)
head(fshdmres$enrichment_result,20)
fsdhm_up <- subset(fshdmres$enrichment_result,p.adjustANOVA<0.02 & s.dist > 0)$set
fshdm_dn <- subset(fshdmres$enrichment_result,p.adjustANOVA<0.02 & s.dist < 0)$set

hmm <- mitch_import( h,DEtype="edger",geneTable=gth )
hres <- mitch_calc(x=hmm,genesets=genesets,priority="effect",resrows=100)
head(hres$enrichment_result,20)
hres_up <- subset(hres$enrichment_result,p.adjustANOVA<0.02 & s.dist > 0)$set
hres_dn <- subset(hres$enrichment_result,p.adjustANOVA<0.02 & s.dist < 0)$set

jmm <- mitch_import( jm,DEtype="edger",geneTable=gtm )
jres <- mitch_calc(x=jmm,genesets=genesets,priority="effect",resrows=100)
head(jres$enrichment_result,20)
jres_up <- subset(jres$enrichment_result,p.adjustANOVA<0.02 & s.dist > 0)$set
jres_dn <- subset(jres$enrichment_result,p.adjustANOVA<0.02 & s.dist < 0)$set

msm <- mitch_import( ms,DEtype="edger",geneTable=gtm )
mres <- mitch_calc(x=msm,genesets=genesets,priority="effect",resrows=100)
head(mres$enrichment_result,20)
mres_up <- subset(mres$enrichment_result,p.adjustANOVA<0.02 & s.dist > 0)$set
mres_dn <- subset(mres$enrichment_result,p.adjustANOVA<0.02 & s.dist < 0)$set

v1 <- list("fshd up"=fsdhm_up,
  "fshd dn"=fshdm_dn,
  "h up"=hres_up,
  "h dn"=hres_dn)

plot(euler(v1),quantities = TRUE)

v1 <- list("fshd up"=fsdhm_up,
  "fshd dn"=fshdm_dn,
  "j up"=jres_up,
  "j dn"=jres_dn)

plot(euler(v1),quantities = TRUE)

v1 <- list("fshd up"=fsdhm_up,
  "fshd dn"=fshdm_dn,
  "m up"=mres_up,
  "m dn"=mres_dn)

plot(euler(v1),quantities = TRUE)

```

Here we run multivariate mitch with different prioritisation schemes.

```{r,mitch,fig.height=8,fig.width=8}

# effect size

res1 <- mitch_calc(x=mg,genesets=genesets,priority="effect",resrows=100)

x <- res1$enrichment_result[1:100,c(4:7)]
rownames(x) <- res1$enrichment_result$set[1:nrow(x)]

colfunc <- colorRampPalette(c("blue","white", "red"))

heatmap.2(as.matrix(x),col=colfunc(30),margins=c(5,25),trace="none",
  scale="none",cexCol=0.6,cexRow=0.25)

unlink("mitch_plots_simple_eff.html")

capture.output(
    mitch_report(res1,"mitch_plots_simple_eff.html")
  ,file = "/dev/null", append = FALSE,
  type = c("output", "message"), split = FALSE)

# significance
res3 <- mitch_calc(x=mg,genesets=genesets,priority="significance",resrows=100)

x <- res3$enrichment_result[1:100,c(4:7)]
rownames(x) <- res3$enrichment_result$set[1:nrow(x)]

colfunc <- colorRampPalette(c("blue","white", "red"))

heatmap.2(as.matrix(x),col=colfunc(30),margins=c(5,25),trace="none",
  scale="none",cexCol=0.6, cexRow=0.25)

# discordance
res2 <- mitch_calc(x=mg,genesets=genesets,priority="SD",resrows=100)

x <- res2$enrichment_result[1:100,c(4:7)]
rownames(x) <- res2$enrichment_result$set[1:nrow(x)]

colfunc <- colorRampPalette(c("blue","white", "red"))

heatmap.2(as.matrix(x),col=colfunc(30),margins=c(5,25),trace="none",
  scale="none",cexCol=0.6, cexRow=0.25)

```

# Session information

```{r,session}

sessionInfo()

```
